Comparison of large scale citizen science data and long term study data for phenology modeling

### 2018-01-26 manuscript/main.tex
first draft sent to coauthors

\author[1]{Shawn D. Taylor}
\author[2]{Ethan P. White}
\affil[1]{UFL}
\affil[2]{UFL}
\corrauthor[1]{Shawn D. Taylor}{shawntaylor@weecology.org}

@@

Plant phenology affects many fields of ecological research such as ecosystem function, carbon dynamics, pollination, and community ecology \citep{richardson2013, cleland2007, tang2016}. Robust phenology models are necessary for forecasting the future state of the climate as well as the impacts on plant and animal communities. At the largest scale variation in the timing of spring leaf out and fall senescence account for a significant amount of error in the carbon budget of earth system models, which has implications for correctly accounting for biosphere-atmosphere feedbacks in long term forecasts \citep{richardson2012}. At local scales, because of species specific responses to temperature and precipitation, changes in phenology can alter the flower community available to pollinators and seed predators throughout the growing season, potentially causing a cascade of affects on abundance, species richness, and interspecific interactions. \citep{diez2012, caradonna2014, ogilvie2017, theobald2017}.

Many phenological studies in the past have relied upon datasets collected at a single location over a long time-frame by a single group \citep{cook2012, roberts2015, iler2013, wolkovich2012}. These datasets have high temporal coverage, with some approaching several decades, and represent the species and climate variability at their respective sites extremely well. Yet it is common for phenology models built with observations from a single site to not transfer well to other sites \citep{garcia-mozo2008, xu2013, olsson2014, basler2016}. Phenology models are species specific, thus this lack of transferability across space can be driven by, among other factors, plasticity in phenology requirements, local adaptation, microclimates, plant age, or population density \citep{kramer1995,diez2012}. Models fit to only a single site can easily be overfit, and being able to generalize across a species range will be important for improving phenology forecasts \citep{richardson2013}. To accomplish this phenology observations will be needed from as much of the full range of species as possible.

@@

Data from long term studies have been invaluable for phenology research for several decades, but datasets with a larger spatial extent and more complete species pool will be needed to model the complex phenology dynamics of ecosystems \citep{richardson2012, diez2012, caradonna2014}. Such datasets from China and Europe have already contributed considerably to phenological research \citep{olsson2014, basler2016, xu2013, zhang2017}, and the NPN dataset has the potential to meet these needs for North America. To date no study has assessed how well the NPN dataset compares to long term study phenology datasets. Here we will compare process based phenology models built using only the NPN data with models built with data from Long Term Study (LTS) datasets. The specific research questions are:
\begin{enumerate}
\item Do parameter estimates of species level models differ when they are built with the NPN dataset versus an LTS dataset?
\item Do estimates of phenological events differ between the NPN and LTS derived models?
\item How well do models built with the NPN dataset transfer to LTS observations, and vice verse?
\end{enumerate}

@@

For each species/phenophase in each of the five datasets we fit a range of phenology models from a naive mean day of year estimate to a more complex two phase forcing model (Table 2). With the exception of two models (naive and linear) the general model form predicts that a phenological event will happen when sufficient forcing units, \(F^{*}\), accumulate from a particular start day of the year. The day forcing accumulation begins, $t_{1}$, can either be estimated or fixed. For the growing degree day model (GDD) forcing units are the degrees in Celsius above an estimated threshold. The Fixed GDD model uses the same form but has fixed values for start day (Jan 1) and temperature threshold (0C). The Alternating model has a variable number of required forcing units defined as function of the total number of days below 0C (NCD). The Uniforc model is similar to the GDD model but has forcing units derived a sigmoid function.

The naive model uses only the mean DOY from all prior observations as the estimated DOY. The linear model uses a regression with the mean spring (Jan. - March) temperature as the only independent variable and DOY as the response variable. Finally two models attempt to explain the spatial variation in phenological requirements. The first spatial model, M1, is an extension of the GDD model which adds a correction in the required forcing using the photoperiod ($L$). The second, the Macroscale Species-specific Budburst model (MSB), uses the mean spring temperature as a linear correction on the total forcing required in the Alternating model. Since there is little to no spatial variation in the LTS datasets, the two spatial models were fit using only the NPN dataset. The resulting parameters, estimates, and errors for the NPN derived M1 and MSB models were compared to the LTS GDD and Alternating models, respectively, for each species and phenophase.

@@

As described above, for each species and phenophase two sets of models were produced: one set of models parameterized using only NPN data, and one set parameterized using only LTS data (with the exception of the M1 and MSB models, see above). The distribution for each parameter in each species/phenophase model was then compared between the two sets using a Mann-Whitney test. This was chosen due to the majority of parameters having non-normal distributions derived from the bootstrapped estimates.

@@

For each species and phenophase this was calculated for all combinations of LTS and NPN derived models, the six phenology models, and the two scales described above.

TODO: Pairwise model error comparison

$$ RMSE_{difference} = \widehat{RMSE_{NPN,s}} - RMSE_{LTS,s} $$

for all $$S = s_{1},...,s_{39}$$ species

@@

\begin{figure}
	\centering
		\includegraphics[width=1\textwidth]{fig_2_estimate_compare.png}
	\caption{Difference between NPN and LTS derived estimates. Each point is, for a single species/phenophase, the root mean square difference between LTS and NPN derived estimates for a specified model. A shows the RMSD over the extent of the NPN dataset, B shows the RMSD at the single sites represented by the LTS datasets.}
\end{figure}

@@

Figure 3 shows the errors for all different model types when tested against the held out data. Models predicting NPN observations (Fig. 3a) have higher error rates than predictions for LTS observations (Fig. 3b). For the former models derived from NPN data made the best predictions, while LTS derived models performed best for the latter. Figure 4 shows the pairwise comparison of LTS and NPN errors. Similar to Figure 3, the NPN models performed best on NPN observations and LTS models performed best on LTS observations.

@@

TODO: figure 4 results

@@

Despite having statistically different parameters to LTS derived models, many NPN derived models were able to produce similar estimates for phenological events. This is (likely) due to models having a non-linear parameter space and being non-identifiable. For example two GDD models with parameters of $t_{1}$=1, $F$=10, $T*$=0 and $t_{1}$=5, $F$=5, $T*$=0 can produce nearly identical results in many, but not all, scenarios. As model complexity increases the ability to produce similar estimates increases. Indeed, the Uniforic models with NPN derived parameters still had estimates within 10 days of LTS derived parameters for most species (Fig. 2) despite having very different parameters (Fig. 1).

@@

Even with similar parameters, the Fixed GDD model was not the best performing model when tested on out of sample data. While the values of additional parameters may be incorrect in the above sense, they nonetheless provide more predictive power. Across the two scales the GDD, Uniforc, and Alternating model were generally the best performing (Fig. 3). Considering only NPN derived models predicting NPN observations the GDD and Uniforc models were the best performing across all species and phenophases, and commonly had nearly identical error values (Fig S1). One goal of ecological forecasting is to inform decision makers about the future state of systems \citep{dietz2017}. Thus the best performing model should be used in these cases to give the best possible forecast. The GDD and Uniforc models have three and four parameters, respectively, therefore the simpler GDD model shows promise for building models with the NPN dataset and making forecasts across large scales.

@@

The models which incorporate a spatial component did not improve over their base models. Correspondence between parameter estimates, estimates of phenological events, and out of sample error rates for the MSB and M1 models were essentially the same as the Alternating and GDD models, respectively. The MSB and M1 models had the lowest error in 9 out of 39 occasions, but never by more than a single day over other models (Fig S1). Spatial variation in phenological requirements is known to exist in plants \citep{zhang2017}, yet here models which attempt to account for this variation did not improve estimates overall. This could be caused either by models not adequately capturing the process driving the spatial variation, the NPN dataset having biases from variation in sampling effort and/or spatial auto-correlation, or some combination of these factors. \cite{basler2016} used the M1 model to predict budburst on six species across Europe and found it was usually among the best models in terms of RMSE, albeit never by more than a single day. Their result was strengthened by a 40 year timeseries across a large region.

>>>>

\author[1]{Shawn D. Taylor}
\author[2]{Joan M. Meiners}
\author[3]{Michael Orr}
\author[4]{Kristina Riemer}
\author[5]{Ethan P. White}

\affil[1,2]{UFL SNRE}
\affil[3]{???}
\affil[4,5]{UFL WEC}
\affil[5]{UFL Bioinformatics}
\corrauthor[1]{Shawn D. Taylor}{shawntaylor@weecology.org}

@@

Plant phenology affects many fields of ecological research such as ecosystem function, carbon dynamics, pollination, and community ecology \citep{richardson2013, cleland2007, tang2016}. Robust plant phenology models are necessary for forecasting the future state of the climate as well as the impacts on plant and animal communities. At the largest scale variation in the timing of spring leaf out and fall senescence account for a significant amount of error in the carbon budget of earth system models, which has implications for correctly accounting for biosphere-atmosphere feedbacks in long term forecasts \citep{richardson2012}. At local scales, because of species specific responses to temperature and precipitation, changes in phenology can alter the flower community available to pollinators and seed predators throughout the growing season, potentially causing a cascade of affects on abundance, species richness, and interspecific interactions. \citep{diez2012, caradonna2014, ogilvie2017, theobald2017}.

Many plant phenological studies in the past have relied upon datasets collected at a single location over a long time-frame by a single group \citep{cook2012, roberts2015, iler2013, wolkovich2012}. These datasets have high temporal coverage, with some approaching several decades, and represent the species and climate variability at their respective sites extremely well. Yet it is common for phenology models built with observations from a single site to not transfer well to other sites \citep{garcia-mozo2008, xu2013, olsson2014, basler2016}. Phenology models are species specific, thus this lack of transferability across space can be driven by, among other factors, plasticity in phenology requirements, local adaptation, microclimates, plant age, or population density \citep{kramer1995,diez2012}. Models fit to only a single site can easily be overfit, and being able to generalize across a species range will be important for improving phenology forecasts \citep{richardson2013}. To accomplish this phenology observations will be needed from as much of the full range of species as possible.

@@

Data from long term studies have been invaluable for phenology research for several decades, but datasets with a larger spatial extent and more complete species pool will be needed to model the complex phenology dynamics of ecosystems \citep{richardson2012, diez2012, caradonna2014}. Such datasets from China and Europe have already contributed considerably to phenological research \citep{olsson2014, basler2016, xu2013, zhang2017}, and the NPN dataset has the potential to meet these needs for North America plant species and communities. To date no study has assessed how well the NPN dataset compares to long term study phenology datasets. Here we will compare process based phenology models built using only the NPN data with models built with data from Long Term Study (LTS) datasets. The specific research questions are:
\begin{enumerate}
\item Do parameter estimates of species level models differ when they are built with the NPN dataset versus an LTS dataset?
\item Do estimates of phenological events differ between the NPN and LTS derived models?
\item How well do models built with the NPN dataset transfer to LTS observations, and vice verse?
\end{enumerate}

@@

For each species/phenophase in each of the five datasets we fit a range of phenology models from a naive mean day of year estimate to a more complex two phase forcing model (Table 2). With the exception of two models (naive and linear) the general model form predicts that a phenological event will happen when sufficient forcing units, \(F^{*}\), accumulate from a particular start day of the year. The day forcing accumulation begins, $t_{1}$, can either be estimated or fixed. For the growing degree day model (GDD) forcing units are the degrees in Celsius above an estimated threshold. The Fixed GDD model uses the same form but has fixed values for start day (Jan 1) and temperature threshold (0C). The Alternating model has a variable number of required forcing units defined as function of the total number of days below 0C (NCD). The Uniforc model is similar to the GDD model but has forcing units derived a sigmoid function. 
 \citep{chuine2000}
The naive model uses only the mean DOY from all prior observations as the estimated DOY. The linear model uses a regression with the mean spring (Jan. - March) temperature as the only independent variable and DOY as the response variable. Finally two models attempt to explain the spatial variation in phenological requirements. The first spatial model, M1, is an extension of the GDD model which adds a correction in the required forcing using the photoperiod ($L$) \citep{blumel2012}. The second, the Macroscale Species-specific Budburst model (MSB), uses the mean spring temperature as a linear correction on the total forcing required in the Alternating model \citep{jeong2013}. Since there is little to no spatial variation in the LTS datasets, the two spatial models were fit using only the NPN dataset. The resulting parameters, estimates, and errors for the NPN derived M1 and MSB models were compared to the LTS GDD and Alternating models, respectively, for each species and phenophase.

@@

As described above, for each species and phenophase two sets of models were produced: one set of models parameterized using only NPN data, and one set parameterized using only LTS data (with the exception of the M1 and MSB models, see above). The distribution for each parameter in each species/phenophase model was compared between the two sets using a Mann-Whitney test. This was chosen due to the majority of parameters having non-normal distributions derived from the bootstrapped estimates.

@@

For each species and phenophase this was calculated for all combinations of LTS and NPN derived models, the eight phenology models, and the two scales described above.

Finally we explored which set of models, either LTS or NPN derived, performed better overall at each scale. We calculated the pairwise difference of RMSE between LTS and NPN derived models for each model type, species, and phenophase at the two different scales. We plotted the density function

$$ f(x) = RMSE_{NPN} - RMSE_{LTS} $$

for each species and phenophases  $$\{ f(x) | s_{1},...,s_{39}\} $$

for each model type and scale.

@@

\begin{figure}
	\centering
		\includegraphics[width=1\textwidth]{fig_2_estimate_compare.png}
	\caption{Difference between NPN and LTS derived estimates. Each point is, for a single species/phenophase, the root mean square difference between LTS and NPN derived estimates for a specified model. A shows the RMSD over the extent of the NPN dataset (a large scale), B shows the RMSD at the single sites represented by the LTS datasets (a small scale).}
\end{figure}

@@

Figure 3 shows the errors for all different model types using the held out data. Models predicting NPN observations (Fig. 3a) have higher error rates than predictions for LTS observations (Fig. 3b). For the former models derived from NPN data made the best predictions, while LTS derived models performed best for the latter. Figure 4 shows the pairwise comparison of LTS and NPN errors. Similar to Figure 3, the NPN models performed best on NPN observations and LTS models performed best on LTS observations.

@@

Figure 4 shows the pairwise comparison of specific species/phenophase combinations for held out data. Overall NPN derived models performed best on held out NPN observations, while LTS derived models performed best on held out LTS observations. The naive and linear models had the largest differences between the two models sets, while the Fixed GDD model had relatively similar errors when evaluated on both NPN and LTS held out observations.

@@

Despite having statistically different parameters to LTS derived models, many NPN derived models were able to produce similar estimates for phenological events. This is likely due to models having a non-linear parameter space and being non-identifiable. For example two GDD models with parameters of $t_{1}$=1, $F$=10, $T*$=0 and $t_{1}$=5, $F$=5, $T*$=0 can produce nearly identical results in many, but not all, scenarios. As model complexity increases the ability to produce similar estimates increases. Indeed, the Uniforic models with NPN derived parameters still had estimates within 10 days of LTS derived parameters for most species (Fig. 2) despite having very different parameters (Fig. 1).

@@

Even with similar parameters, the Fixed GDD model was not the best performing model when tested on out of sample data. While the values of additional parameters may be incorrect in the above sense, they nonetheless provide more predictive power. Across the two scales the GDD, Uniforc, and Alternating model were generally the best performing (Fig. 3). Considering only NPN derived models predicting NPN observations the GDD and Uniforc models were the best performing across all species and phenophases, and commonly had nearly identical error values (Fig S1). One goal of ecological forecasting is to inform decision makers about the future state of systems \citep{dietz2018}. Thus the best performing model should be used in these cases to give the best possible forecast. The GDD and Uniforc models show promise for building models with the NPN dataset and making plant phenology forecasts across large scales.

@@

The models which incorporate a spatial component did not improve over their base models. Correspondence between parameter estimates (Fig. 1), estimates of phenological events (Fig. 2), and out of sample error rates (Fig. 3 & 4) for the MSB and M1 models were essentially the same as the Alternating and GDD models, respectively. The MSB and M1 models had the lowest error in 9 out of 39 occasions, but never by more than a single day over other models (Fig S1). Spatial variation in phenological requirements is known to exist in plants \citep{zhang2017}, yet here models which attempt to account for this variation did not improve estimates overall. This could be caused either by models not adequately capturing the process driving the spatial variation, the NPN dataset having biases from variation in sampling effort and/or spatial auto-correlation, or some combination of these factors. \cite{basler2016} used the M1 model to predict budburst on six species across Europe and found it was usually among the best models in terms of RMSE, albeit never by more than a single day. Their result was strengthened by a 40 year timeseries across a large region. \cite{chuine2017} noted that incorporating the spatial vaiation in warming requirements as a key issue in future phenology research. We feel the NPN dataset, as well as other large scale phenology datasets, will be key in addressing this and other phenological research needs.
